Comparing probabilistic predictive models applied to football

### 2015-08-16 template.tex
revisao rafael

% Abstract (Do not use inserted blank lines, i.e. \\)
\abstract{We create three simple multinomial-Dirichlet models and compare them to two well-known sophisticated statistical models for association football (soccer) predictions according to their performance in predicting the full-time results for the second round of the 2014 Brazilian first division championship. The predictive models are compared using three proper scoring rules, the proportion of errors and the calibration assessment. Our results show similar predictive performances for all the five methods, thus discouraging the use of overcomplicated models for match result forecast.}

@@

    Bothered by such concerns, our goal was to answer the following question: is it possible, using a simple probabilistic model, to predict results of football matches better than more sophisticated or complex models?
    To answer this, we compared two sophisticated models that consider multiple covariates with simple multinomial-Dirichlet models that consider only the number of matches won, tied or lost by each team as inputs.
    The models were tested for the second round of the Brazilian football championship of first division and compared using standard metrics or scoring rules.
    We also used other criteria such as the proportion of matches that were predicted ``correctly'' by each model and a measure of calibration, both explained in detail below.
    According to the adopted criteria, the simple models showed a similar predictive performance to the complex ones.

    The predictions of the two benchmark models were published on Internet websites before every matchday, and are treated by us as black-box models (BB1 and BB2).
    Since one of the authors is responsible for one of these models and the other was described in a dissertation \citep{arruda2000}, we know that both models are based on Holgate bivariate Poisson regression considering as explanatory variables the home field advantage, attack and defense strength of each team, and provide the probabilities of all possible outcomes (home team wins, draws or loses) for each match.
    Because both models are based on elaborate parametric assumptions, they are suitable for the comparison we are proposing: simple versus complex models.

    To be precise, we would have to define what is a ``simple'' model and what is a ``complex'' model. Based on the opening quotations, one may say that the number of parameters defines the complexity of a model, which is not exactly the definition we will use, since we consider other criteria, like the interpretation of explanatory variables and mathematical structure of the statistical model.

@@

    Although our benchmark models were here defined as complex, we can find much more elaborate models in the literature, hindering the estimation and prediction processes.
    Among them, we can cite those based on modeling the match as a stochastic process evolving through time \citep{Dixon98, Volf2009, Titman2015}, those allowing for the team performance parameters to change along the season \citep{Rue2000, Crowder2002, Owen2011, Koopman2015} and those modeling dependence between number of goals by means of bivariate count distributions \citep{Dixon97, Karlis2003, McHale2007, McHale2011}. Models that predict the number of goals have more specific objectives than those prediciting the final outcomes and therefore require more elaborate modeling, justifying the complexity of such models.

    According to standard scoring rules and other criteria here adopted to qualify the predictions, our findings show that the predictive power of the three simple models was quite similar to the benchmark models.
    %We compared the probabilistic predictions of three simple models with the ones provided by two more complex models for the second round of the 2014 Brazilian football championships of first and second divisions.
    {\color{red}This work thus indicates that while there has been a boom in the literature with the proposal of several fancy and involved models, especially in the era of big data, simple models can many times predict the final outcome as well as their complex counterparts. Nevertheless, the complex models cited here are able to answer different questions besides the final outcome that the simple multinomial models are not, for instance, they can estimate teams' performance parameters allowing for ranking teams according to attack and deffensive quality. Thus, we can not assert that complex models are useless.}

    This is often unnoticed by users of such methods, which leads to the overuse of models that are
    hard to interpret and to implement, and that also require one to keep track of several covariates. We would call this phenomenon the {\it king's nudity} { in allusion to Anderson's {\it The Emperor's New Clothes}, \cite{emperor}: ``But he isn't wearing anything at all!''. However, this is not always the case. In this spirit, since we have explored only the prediction of football matches, we would like to study if the same extends to other fields such as, for instance, weather or economics forecasts.

>>>>

% Abstract (Do not use inserted blank lines, i.e. \\)
\abstract{We create three simple multinomial-Dirichlet models and compare them to  two 
well-known sophisticated statistical models for association football (soccer) predictions according to their performance in predicting the full-time results for the second round of the 2014 Brazilian first division championship. The predictive models are compared using three proper scoring rules, the proportion of errors and the calibration assessment. Our results show similar predictive performances for all the five methods, thus {\color{blue} indicating that the new models offer simpler and more interpretable
predictions while maintaining predictive power. This  discourages the use of the two popular models for match result forecast in the Brazilian football championship.}}

@@

	Bothered by such concerns, our goal was to answer the following question: is it possible, using a simple probabilistic model, to predict results of football matches better than more sophisticated or complex models?
	{\color{blue}To answer this, we developed simple multinomial-Dirichlet models that consider only the number of matches won, tied or lost by each team as inputs. We then compared them with the two most popular 
	 models used to predict the outcomes of the Brazilian football championship. These two sophisticated models are widely used by football fans, and  they consider multiple covariates as inputs.}
	The models were tested for the second round of the Brazilian football championship of first division and compared using standard metrics or scoring rules. 
	We also used other criteria such as the proportion of matches that were predicted ``correctly'' by each model and a measure of calibration, both explained in detail below.
	According to the adopted criteria, the simple models showed a similar predictive performance to the complex ones.

	The predictions of the two benchmark models are published on Internet websites before every matchday, and are treated by us as black-box models (BB1 and BB2).
	Since one of the authors is responsible for one of these models and the other was described in a dissertation \citep{arruda2000}, we know that both models are based on Holgate bivariate Poisson regression considering as explanatory variables the home field advantage, attack and defense strength of each team, and provide the probabilities of all possible outcomes (home team wins, draws or loses) for each match.
{\color{red}sugiro tirar - Because both models are based on elaborate parametric assumptions, they are suitable for the comparison we are proposing: simple versus complex models. 
	}

	{\color{red}sugiro tirar -To be precise, we would have to define what is a ``simple'' model and what is a ``complex'' model. Based on the opening quotations, one may say that the number of parameters defines the complexity of a model, which is not exactly the definition we will use, since we consider other criteria, like the interpretation of explanatory variables and mathematical structure of the statistical model.}

@@

	{\color{blue}The benchmark models used in this work were chosen because of their wide popularity among football fans in Brazil.
	Several other models are also available in the literature, some of which are even more complex.
	Among them, we can cite those based on modeling the match as a stochastic process evolving through time \citep{Dixon98, Volf2009, Titman2015}, those allowing for the team performance parameters to change along the season \citep{Rue2000, Crowder2002, Owen2011, Koopman2015} and those modeling dependence between number of goals by means of bivariate count distributions \citep{Dixon97, Karlis2003, McHale2007, McHale2011}. 
Contrary to the simple multinomial models we propose,	some of these approaches are able to answer several questions;  for instance, they can estimate teams' performance parameters allowing for ranking teams according to attack and deffensive quality, and 
 they can also predict the final outcomes of matches. We therefore do not claim that our models should replace
	current state-of-the-art approaches. Instead, this work
	indicates that while there has been a boom in the literature with the proposal of several fancy and involved models,
	simple models can many times predict the final outcome as well as their complex counterparts. 
	The importance of such finding is directly related to the model's usability in practice: creators of the
	benchmark models used in this work often say a difficulty they face is that
	wrong predictions often  generate anger in football fans, which is then translated into distrust
	in subsequent predictions. Because of the complexity of those models, they find it hard to explain predicted outcomes
	to users. This is where using simple models pays off:  the multinomial models yield results that are
	easy to explain; they simply involve counts of losses, wins and draws!
	They allow us to offer simple explanations to the football fans about the predictions that were made.}

{\color{blue}The fact that simple models are often as effective as their competitors is often unnoticed by users of such methods, which leads to the overuse of models that are often
	hard to interpret and to implement, and that also require one to keep track of several covariates. We would call this phenomenon the {\it king's nudity}  in allusion to Anderson's {\it The Emperor's New Clothes}, \cite{emperor}: ``But he isn't wearing anything at all!''. In this spirit, since we have explored only the prediction of football matches in the specific case o Brazilian football championship, we would like to study if the same extends to other fields such as, for instance, weather or economics forecasts.}
